---
title: "1 Install R/R Studio"
---

## 🎯Learning goals

After working through Tutorial 1, you'll...

-   know how to install R and R Studio
-   understand the main set-up of R Studio

## Object- and function-oriented programming

First things first: R is an object- and function-oriented programming language. Chambers [(2014, p. 4)](https://arxiv.org/pdf/1409.3531.pdf) has nicely summarized what this means:

-   Everything that exists is an object.
-   Everything that happens is a function call.

We assign values (for instance, single numbers/letters, several numbers/letters, or whole data sets) to objects in R to work with them/do computations. For instance, the following command in R would assign the word "hello" to the object *word* by using the `<-` sign (a function used for assigning values to objects):

```{r,hello-0, eval = TRUE, echo = TRUE}
word <- "hello"
```

Objects have specific properties that determine which types of calculations can be done with them (and which cannot). For instance, the object *word* is characterized by the fact that it consists of characters (i.e., a word) - which may, for instance, prohibit you to calculate the mean of this object (which is something only possible for objects consisting of numeric data).

This already hints towards the second, important aspect of R: It is influenced by functional programming, meaning that everything we *do* in R is a function call. R uses functions to assign specific values (for instance, a single number or word, several numbers or words, whole data sets) to objects. In short, what you will learn by learning programming in R is how to write functions for making R do the calculations you need.

Let's take SPSS - which, I assume, many of you have worked with previously - as a comparison. If you import, edit, or analyze a data set in SPSS, you'll use the click-and-point menu to change variable values, calculate variables' means, or export data.

R works differently: You assign the data set to an object - for instance an object called "word", as done previously. This assignment enables you to work with the data: You can now call specific functions to work with or change this object. For instance, if you want to add another word to the object *word*, for instance "and good morning", you could do that by using a specific function called `paste0()`, which takes the original object *word* and adds your additional words " and good morning" to overwrite the old object *word*:

```{r,hello-0b, eval = TRUE, echo = TRUE}
word <- paste0(word, " and good morning")
word
```

Different to using the click-and-point menu in SPSS, you thus **need to write your own code** to import, edit, or analyze any type of data in R. Luckily, R already includes a lot of predefined functions meaning that we do not have write all of these functions ourselves.

## Base-R vs. tidyverse & coding style

There are two "ways" of using R: via Base-R or via the [tidyverse](https://www.tidyverse.org/), [see this overview for a comparison](https://towardsdatascience.com/tidyverse-vs-base-r-how-to-choose-the-best-framework-for-you-29b702bdb384).

Base-R includes all functions that come with initially installing R. A more recent development is the tidyverse, a collection of additional packages (and coding conventions) that is especially suited for beginners. Tidyverse was developed by Hadley Wickham and his team at RStudio (the interface we use for programming with R).

Best practice when coding with R is to either stick to Base-R or the tidyverse, as this makes your code more streamlined and readabile. In this class, we will rely on the tidyverse. You will learn a little bit of Base-R in tutorials 2-3 to know how to read this code (e.g., when stumbling across online forums when looking for R-related code solutions). However, we will then rely on tidyverse in subsequent tutorials.

This also means that I will use the [tidyverse coding style](https://style.tidyverse.org/). With this coding style, we make sure to consistently name things (e.g., with lower letters), use blank spaces, where needed. In short, it makes our code more readable.

If your're interested in this, see a nice summary by IfKW scholar Lara Kobilke [here](https://lkobilke.github.io/introduction_to_R/tutorial-good-coding-style-project-management.html).

## 
